### 2017-09-28 paper/sexSelectionSpacing-ver1.tex
Add draft of discussion of how to compare survival curves

Spacing between births has long been used as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference was therefore often associated with shorter average spacing 
after the births of girls than boys, which, in turn, was associated with worse 
health outcomes for girls.%
\footnote{
See, for example, 
\citet{Das1987}, \citet{Rahman1993}, \citet{Pong1994}, \citet{Haughton1996},
\citet{Arnold1997}, and \citet{Soest2012} on shorter birth spacing after
girls
and 
\citet{arnold98}, \citet{Whitworth2002}, \citet{Rutstein2005},
and \citet{Conde-Agudelo2006} on the association between shorter birth
spacing and worse health outcomes for girls.
Parents were also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Altindag2016}
}

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship between 
son preference and birth spacing.
This is because each sex-selective abortion increases the duration between 
observed births by approximately a year.%
\footnote{
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}
As a result, we now have a situation where families with the \emph{strongest} son 
preference likely show the \emph{longest} spacing after the birth of a daughter.
But, to further complicate matters, we may still observe short spacing 
after the birth of daughter as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

@@

In this paper, I introduce and apply a novel method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the last 
birth to affect the decision on sex selection.
For example, as the duration from last birth becomes long enough, parents 
may reverse their decision to use prenatal sex determination,
and carry the next pregnancy to term whether male or female.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

@@

>>>>

Spacing between births has long been used as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference was therefore often associated with shorter spacing after the 
births of girls than boys, which, in turn, was associated with worse 
health outcomes for girls.%
\footnote{
See, for example, 
\citet{Das1987}, \citet{Rahman1993}, \citet{Pong1994}, \citet{Haughton1996},
\citet{Arnold1997}, and \citet{Soest2012} on shorter birth spacing after
girls
and 
\citet{arnold98}, \citet{Whitworth2002}, \citet{Rutstein2005},
and \citet{Conde-Agudelo2006} on the association between shorter birth
spacing and worse health outcomes for girls.
Parents were also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Altindag2016}
}

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship between 
son preference and birth spacing.
This is because each sex-selective abortion substantially increases the 
spacing from last birth. 
For every abortion a woman has we should expect the duration from last
birth to increase by approximately a year.%
\footnote{
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

As a result, we now have a situation where families with the \emph{strongest} son 
preference may show the \emph{longest} spacing after the birth of a daughter
because of the use of sex-selective abortions.
But, to further complicate matters, we may still observe short spacing 
after the birth of daughter as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

@@

In this paper, I introduce and apply a novel method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the last 
birth to affect the decision on sex selection.
For example, as the duration from last birth becomes long enough, even 
parents with strong son preference may reverse their decision to use 
prenatal sex determination, and carry the next pregnancy to term whether 
male or female.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

@@

[How to compare survival curves?]

In principle, in order to understand how birth spacing changes in response
to the introduction of prenatal sex determination one could simply compare survival 
curves between different sex compositions of prior births over time.
If the idea that the availability of sex-selective abortion increases the
spacing more after the birth of girls than after the birth of boys is 
correct, we should see the two survival curves become closer to each other
and we might even be able to directly observe that the survival curve for prior
child being a girl is above the survival curve for prior child being a boy,
instead of substantially below as was generally the case before prenatal
sex determination become widely available.

There are, however, two (interrelated) problems.
First, the probability that a woman will progress to a next birth  
is not the same across different sex compositions of prior children.
Specifically, one of the ways that past research has tried to establish
son preference in fertility behavior is to compare parity progression
rates for different sex compositions [TK references - see footnote 1].
The idea was that if there is son preference a couple would be less likely
to progress to a next birth if they have one or more son already for a
given number of prior births.

The fact that the probability of having a next birth also depends on the
sex composition complicates the comparison of survival curves. 
A lower parity progression probability will naturally make the survival
curve flatter.
This means that the survival curve for a girl composition will be more
likely to be below the boy composition.
Even if the use of sex selection is substantial enough to move the
survival curve for girl composition to the right of the boy
composition the lower parity progression probability for boy 
composition means that the survival curves would eventually have to
cross.
This makes it difficult---if not impossible---to design a direct test
of how the survival curves compare.

The problem of different parity progression rates across different
sex composition is exacerbated by falling fertility over time.
This is because there are now more parities where they may be a 
substantial differences in parity progression rates and because
the difference in parity progression for a given parity will 
become larger if there is son preference.
As the parity progression probability decreases it becomes impossible
to capture the time until a certain survival point (like 0.5 of women
having had a child; the survival curve may never actually reach this
point).

In addition, it is possible that falling fertility also generically 
leads to longer spacing between births.
This means that we have to separate out which part of the longer
spacing comes from falling fertility (and therefore should also
affect boy composition) and which parts comes from increased use
of sex selection.

[We should expect sex selection to be mainly used for the last---or close to
last---birth that a couple expect to have because of the cost and risk
of the procedures.]

One solution to this problem is to condition on the parity progression
rates.
That is, instead of comparing say how long it takes for 50\% of women
who began the spell to have the next birth across boy and girl composition, 
which would work well if the next birth was nearly universal, we 
could calculate how long it takes for 50\% of those women who will 
eventually have a next child to have that child based on the estimated
survival curves.
This number can then be compared across sex compositions to 
understand if the spell length for girl composition increases
more than for boy composition with the introduction of sex selection.

Obviously, this should be calculated for different percentages.
Actually, it probably makes more sense to show this graphically
instead. 
All compositions would begin at 100\% and end at 0\% because
we condition on having the next child.

Note, this is not the same as simply calculating the average 
\emph{observed} spell lengths among women who have a given parity
child in the survey because that number does not take into account
the censoring of some spells that will eventually lead to a birth.
